Practical Hilbert space approximate Bayesian Gaussian processes for probabilistic programming

### 2022-02-22 Paper/Manuscript_Stats&Comp/manuscript_STCO.tex
Edits to Sec. 4. Mainly typos and minor errors. Unified wording across whole paper as well.

A GP model with a periodic covariance function does no fit in the framework of the HSGP approximation covered in this study as a periodic covariance function has not a spectral representation, but it has also a low-rank representation. In Appendix~\ref{sec_periodic}, we briefly present the approximate linear representation of a periodic squared exponential covariance function as developed by \citet{solin2014explicit}, analyze the accuracy of this approximation and, finally, derive the GP model with this approximate periodic square exponential covariance function.

@@

Ultimately, these recommendations are based on the relationships among the number of basis functions $m$, the boundary condition $L$, and the length-scale $\ell$, which depend on the particular choice of the kernel function. In this work we investigate these relationships for the square exponential and the Mat{\'e}rn ($\nu=3/2$ and $\nu=5/2$) covariance functions in the present section, and for the periodic squared exponential covariance function in Appendix~\ref{sec_periodic}. For other kernels, the relationships will be slightly different depending on the smoothness or wigglyness of the covariance function.

@@

Next, we analyze how the interaction between $m$ and $c$ affects the quality of the approximation. The length-scale and marginal variance of the covariance function will no longer be fixed but instead we compute the joint posterior distribution of the function values and the hyperparameters using the dynamic HMC algorithm implemented in Stan \citep{StanTeam:2021} for both the exact GP and the HSGP models. Figure~\ref{fig3_Post_part1} shows the posterior predictive mean and standard deviation of the function as well as the covariance function obtained after fitting the model for varying $m$ and $c$. Figure \ref{fig4_MSE_vs_J} shows the root mean square error (RMSE) of the HSGP models computed against the exact GP model. Figure \ref{fig5_lscale_vs_J} shows the estimated length-scale and marginal variance for the exact GP model and the HSGP models. Looking at the RMSEs in Figure \ref{fig4_MSE_vs_J}, we can conclude that the optimal choice in terms of precision and computation time for this example would be $m = 15$ basis functions and a boundary factor between $c = 1.5$ and $c = 2.5$. Further, the less conservative choice of $m = 10$ and $c = 1.5$  could also produce a sufficiently accurately approximation depending on the application. We may also come to the same conclusion by looking at the posterior predictions and covariance function plots in Figure~\ref{fig3_Post_part1}. From these results, some general conclusions may be drawn:

\begin{itemize}
\item As $c$ increases, $m$ has to increase as well (and vice versa). This is consistent with the expression for the eigenvalues in eq.~\eqref{eq_eigenvalue}, where $L$ appears in the denominator.
\item There exists a minimum $c$ below which an accurate approximation will never be achieved regardless of the number of basis functions $m$.
\end{itemize}

\subsection{Near linear proportionality between $m$, $c$, and $l$} \label{subsec_theoretic_linearity}

@@

% There is a clear relation between the number of basis functions $m$ and the boundary factor $c$ with the length-scale $\ell$ of the approximated function. In this section we discuss about the existing theoretical evidence that the minimum number of basis functions $m$ needed for an accurate approximation is linearly proportional to the the boundary factor $c$ and inversely proportional to the lengthscale $l$.
% In the next Section \ref{subsec_empiric_relations} we conduct an empirical analysis through simulations to prove this expected linear proportionality between $m$, $c$ and $l$ and compute the specific coefficients of proportionality for every kernel.

% {\color{blue}Aki: We can also look at the mean zero-level upcrossings, which for QE is
% proportional to 1/l which gives as that part, and the effect of c should also
% be quite linear when c is not very close to 1 and m is not very small as it
% linearly makes the interval shorter so that the there are less zero-level
% upcrossings available in the basis functions. You can visualize this using
% {\small \url{https://avehtari.github.io/casestudies/Motorcycle/motorcycle_gpcourse.html}} and
% look how the plot of 6 basis functions change when you set c=1, c=1.5, c=2.25
% (ie $1.5^2$). Instead of looking at mean zero-level crossings, we can look at the
% spectral densities and when they go to zero. This is illustrated in the same
% notebook after that figure. E.g. with lengthscale l=1,l=1/2, l=1/4, l=1/8 to
% explain 99\% of the prior variance we need 4, 8, 16, and 32 basis functions
% (linear in l) and to explain 80\% of the prior variance we need 2, 4, 8, 16
% basis functions. These spectral densities don't depend on c. When c increaes
% the basis functions are truncated and for example going from c=1.5 to c=2.25,
% basis function 9 has the flexibility basis function 6 had, so 1.5 more basis
% functions are needed. So we have proportionality c/l. }

@@

\subsection{Empirical discovering of the functional form of the relationships between $m$, $c$ and $l$} \label{subsec_empiric_relations}

\begin{figure*}
\centering
\subfigure{\includegraphics[width=\textwidth, trim = 0mm 0mm 0mm 0mm, clip]{fig6_relationships.png}}
\caption{Relation among the minimum number of basis functions $m$, the boundary factor $c$ \hspace{0.1mm} ($c = \frac{L}{S}$) \hspace{0.1mm} and the length-scale normalized by the half-range of the data ($\frac{\ell}{S}$), for the square exponential, Mat\'ern ($\nu$=3/2) and Mat\'ern ($\nu$=5/2) covariance functions.}
  \label{fig6_relationships}
\end{figure*}

Empirical simulations are carried out to analyze the relationships between $m$, $c$ and $l$. Figure \ref{fig6_relationships} depicts how these three factors interact and affect the accuracy of the HSGP approximation for a GP with square exponential, Mat\'ern ($\nu$=3/2) and Mat\'ern ($\nu$=5/2) covariance functions and a single input dimension. More precisely, for a given GP model (with a particular covariance function) with length-scale $\ell$ and given a boundary factor $c$, Figure \ref{fig6_relationships} shows the minimum number of basis functions $m$ required to obtain an accurate approximation in the sense of satisfying eq. (\ref{eq_diff_covs}). We considered an approximation to be a sufficiently accurate when the total variation difference between the approximate and exact covariance functions, $\varepsilon$ in eq. (\ref{eq_diff_covs}), is below 1$\%$ of the total area under the curve of the exact covariance function $k$ such that
%
\begin{equation} \label{eq:rel_total_distance}
 \int | k(\tau) - \tilde{k}_m(\tau)|  \,\mathrm{d}\tau  < 0.01 \int k(\tau) \,\mathrm{d}\tau,
\end{equation}
%
where $\tilde{k}_m$ is the approximate covariance function with $m$ basis functions. Alternatively, these figures can be understood as providing the minimum $c$ that we should use for given $\ell$ and $m$. Of course, we may also read it as providing the minimum $\ell$ that can be approximated with high accuracy given $m$ and $c$. We obtain the following main conclusions:

\begin{itemize}
\item As $\ell$ increases, $m$ required for an accurate approximation decrease.
\item The lower $c$, the smaller $m$ can and $\ell$ must be to achieve an accurate approximation.
\item For a given $\ell$ there exist a minimum $c$ under which a accurate approximation is never going to be achieved regardless of $m$. This fact can be seen in Figure~\ref{fig6_relationships} as the contour lines which represent $c$ have an end in function of $\ell$ (Valid $c$ are restricted in function of $\ell$). As $\ell$ increases, the minimum valid $c$ also increases.
\end{itemize}

@@

As explained in Section~\ref{subsec_theoretic_linearity}, when $c$ is large enough, there is a near
linear proportionality between $m$, $\frac{l}{S}$ (lengthscale $l$ relative to the half range of the input domain $S$) and $c$. To obtain practical numerical functions that can be used to guide the selection of these parameters, we have empirically derived the practically useful constant terms. We require a lower bound for $c$ of $c \geq 1.2$ such that the equations below are precise enough for practical application.

@@

\paragraph*{Squared exponential}:
%
\begin{align}\label{eq_m_l_QE}
&m= 1.75 \, \frac{c}{\nicefrac{l}{S}} \;\; \Leftrightarrow \;\; \nicefrac{l}{S}= 1.75 \, \frac{c}{m},
\end{align}
%
with
%
\begin{align}\label{eq_c_vs_l_QE}
&c \geq \, 3.2 \, \nicefrac{l}{S} \;\; \& \;\, c \geq \, 1.2      
\end{align}

\paragraph*{Mat\'ern ($\nu$=5/2)}:
%
\begin{align}\label{eq_m_l_mat52}
&m= 2.65 \, \frac{c}{\nicefrac{l}{S}} \;\; \Leftrightarrow \;\; \nicefrac{l}{S}= 2.65 \, \frac{c}{m},
\end{align}
%
with
%
\begin{align}\label{eq_c_vs_l_mat52}
&c \, \geq \, 4.1 \, \nicefrac{l}{S} \;\; \& \;\, c \geq \, 1.2       
\end{align}

\paragraph*{Mat\'ern ($\nu$=3/2)}:
%
\begin{align}\label{eq_m_l_mat32}
&m= 3.42 \, \frac{c}{\nicefrac{l}{S}} \;\; \Leftrightarrow \;\; \nicefrac{l}{S}= 3.42 \, \frac{c}{m},
\end{align}
%
with
%
\begin{align}\label{eq_c_vs_l_mat32}
&c \, \geq \, 4.5 \, \nicefrac{l}{S} \;\; \& \;\, c \geq \, 1.2        
\end{align}

These constants vary monotonically with respect to $\nu$ (squared exponential corresponding to Mat\'ern with $\nu \rightarrow \infty$). Using the formula for Mat\'ern ($\nu$=3/2) provides the largest $m$ and $c$, and thus this formula alone could be used as a conservative choice for all Mat\'ern covariance functions with $\nu \geq 3/2$ and likely as a good initial guess for many other covariance functions. If the aim is to find minimal $m$ to speedup the computation, a further refined formula can be obtained for new covariance functions.

Figure~\ref{fig6_relationships} and previous equations (\ref{eq_m_l_QE})-(\ref{eq_c_vs_l_mat52}) were obtained for a GP with a unidimensional covariance function, which result in a surfaces depending on three variables, $m$, $c$ and $\ell$. Equivalent results for a GP model with a two-dimensional covariance function would result in a surface depending on four variables, $m$, $c$, $\ell_1$ and $\ell_2$. More precisely, in the multi-dimensional case, whether the approximation is close enough might depend only on the ratio between wigglyness in every dimensions. For instance, in the two-dimensional case, it would depend on the ratio between $\ell_1$ and $\ell_2$. Future research will focus on building useful graphs or analytical models that provide these relations in multi-dimensional cases. However, as an approximation, we can use the unidimensional GP conclusions in Figure \ref{fig6_relationships} or equations (\ref{eq_m_l_QE})-(\ref{eq_c_vs_l_mat52}) to check the accuracy by analyzing individually the different dimensions of a multidimensional GP model.

@@

The numerical equation governing the relationships between $m$ and $\ell$ for a periodic squared exponential covariance function was estimated in eq.~\ref{eq_J_l_periodic} which is depicted next:
%
\begin{align*}
&m \geq \frac{3.72}{\ell} \;\; \Leftrightarrow \;\; \ell \geq \frac{3.72}{m}
\end{align*}

@@

Guidelines for updating $m$ and/or $c$ are:

\begin{itemize}
	\item $c$ must be big enough for the given length-scale $l$ of the function to be learned and,
	\item $m$ must be big enough for the given $\ell$ and $c$. 
\end{itemize}

@@

If the initial lengthscale is not far from the optimal, and larger than minimal $c$ and $m$ are used in the initial HSGP model, it is likely that the initial results are already sufficiently accurate. The following iterative procedure works as safety approach and also allows finding a smaller sufficient $m$ which can be useful if the user wants to optimize the computational speed for repeated fitting (e.g., in cross-validation or simulation based calibration).

@@

Based on the above proposed diagnostics, we obtain a simple, iterative step-by-step procedure that users can follow in order to obtain an accurate HSGP approximation. The procedure is split into two phases, Phase A and B, which have to be completed consecutively.
%In the following, we will refer to the Equations from Section XXX for the double-exponential kernel, but the same procedure works for other kernels by using the corresponding Equations provided in the same Section.

@@

\item[A1.] Make an initial guess on the lengthscale $l_1$. If there is no useful information available, we recommend to start with a large lengthscale, about $\ell_1 \in [0.5, 1]$. Set the iteration index to $i=1$.

\item[A2.] Obtain the minimum valid boundary factor $c_i$ determined by $\ell = \ell_i$ for the given kernel as per Section \ref{sec_num_equations}.

\item[A3.] Obtain the mimimum valid number of basis functions $m_i$ determined by $\ell = \ell_i$ and $c = c_i$ for the given kernel as per Section \ref{sec_num_equations}.

@@

\item[A5.] Perform the lengthscale diagnostic by checking if $\hat{\ell}_i - 0.01 \geq \ell_i$.

@@

If the diagnostic is TRUE, the HSGP approximation is close to be sufficiently accurate. Continue with Phase B.

@@

\item[B1.] For the current HSGP model, compute measures of predictive accuracy, for example, root mean square error (RMSE), coefficient of determination ($R^2$), or expected log predicitve density (ELPD).

@@

\item[B4.] Obtain the mimimum valid lengthscale $\ell_i$ that can be accurately estimated by $m = m_i$ and $c = c_i$ for the given kernel as per Section \ref{sec_num_equations}.

@@

\item[B6.] Perform the lengthscale diagnostic by checking if $\hat{\ell}_i - 0.01 \geq \ell_i$.

@@

If all the stability checks succeeed, the HSGP approximation of the latest model should be sufficiently accurate and the procedure ends here.

@@

\begin{figure*}
\centering
\subfigure{\includegraphics[width=\textwidth, trim = 0mm 0mm 0mm 0mm, clip]{fig7_lscale_comparison.png}}
\caption{Exact GP length-scale ($\ell_{GP}$) and approximate HSGP lengthscale ($\hat{\ell}$) for various datasets with varying smoothness (left) and their corresponding root mean squared errors (RMSE) computed with the exact GP model as the reference.}
  \label{fig7_lscale_comparison}
\end{figure*}

@@

In Section \ref{sec_accuracy}, we thoroughly studied the relationship between the number of basis functions in the approximation and the approximation accuracy across different configurations. We specifically focused on the Matérn and squared exponential families of covariance functions, yet there exists other families of stationary covariance functions. The basis function approximation can easily be implemented for any stationary covariance function, where the spectral density function is available. The assess the accuracy of a basis function approximation for a kernel, where the diagnostic plots like Figure \ref{fig6_relationships} or equations (\ref{eq_m_l_QE})-(\ref{eq_c_vs_l_mat52}) are not available, we suggest to use the relative total variational distance between the true covariance function and the approximate covariance function as given in eq. \eqref{eq:rel_total_distance}. Ensuring that the relative distance is bounded by a small constant for the relevant lengthscale implies a high quality approximation.

@@

A GP model with a periodic covariance function does no fit in the framework of the HSGP approximation covered in this study, but it has also a low-rank representation. In this section, we first give a brief presentation of the results by \citet{solin2014explicit}, who obtain an approximate linear representation of a periodic squared exponential covariance function based on expanding the periodic covariance function into a series of stochastic resonators. Secondly, we analyze the accuracy of this approximation and, finally, we derive the GP model with this approximate periodic square exponential covariance function.

>>>>

A GP model with a periodic covariance function does no fit in the framework of the HSGP approximation covered in this study as a periodic covariance function has not a spectral representation, but it has also a low-rank representation. In Appendix~\ref{sec_periodic}, we briefly present the approximate linear representation of a periodic squared exponential covariance function as developed by \citet{solin2014explicit}, analyze the accuracy of this approximation and, finally, derive the GP model with this approximate periodic squared exponential covariance function.

@@

Ultimately, these recommendations are based on the relationships among the number of basis functions $m$, the boundary condition $L$, and the length-scale $\ell$, which depend on the particular choice of the kernel function. In this work we investigate these relationships for the squared exponential and the Mat{\'e}rn ($\nu=3/2$ and $\nu=5/2$) covariance functions in the present section, and for the periodic squared exponential covariance function in Appendix~\ref{sec_periodic}. For other kernels, the relationships will be slightly different depending on the smoothness or wigglyness of the covariance function.

@@

Next, we analyze how the interaction between $m$ and $c$ affects the quality of the approximation. The length-scale and marginal variance of the covariance function will no longer be fixed but instead we compute the joint posterior distribution of the function values and the hyperparameters using the dynamic HMC algorithm implemented in Stan \citep{StanTeam:2021} for both the exact GP and the HSGP models. Figure~\ref{fig3_Post_part1} shows the posterior predictive mean and standard deviation of the function as well as the covariance function obtained after fitting the model for varying $m$ and~$c$. Figure~\ref{fig4_MSE_vs_J} shows the root mean square error (RMSE) of the HSGP models computed against the exact GP model. Figure~\ref{fig5_lscale_vs_J} shows the estimated length-scale and marginal variance for the exact GP model and the HSGP models. Looking at the RMSEs in Figure~\ref{fig4_MSE_vs_J}, we can conclude that the optimal choice in terms of precision and computation time for this example would be $m = 15$ basis functions and a boundary factor between $c = 1.5$ and $c = 2.5$. Further, the less conservative choice of $m = 10$ and $c = 1.5$ could also produce a sufficiently accurately approximation depending on the application. We may also come to the same conclusion by looking at the posterior predictions and covariance function plots in Figure~\ref{fig3_Post_part1}. From these results, some general conclusions may be drawn:
%
\begin{itemize}
\item As $c$ increases, $m$ has to increase as well (and vice versa). This is consistent with the expression for the eigenvalues in eq.~\eqref{eq_eigenvalue}, where $L$ appears in the denominator.
\item There exists a minimum $c$ below which an accurate approximation will never be achieved regardless of the number of basis functions $m$.
\end{itemize}

\subsection{Near linear proportionality between $m$, $c$, and $\ell$} \label{subsec_theoretic_linearity}

@@

% There is a clear relation between the number of basis functions $m$ and the boundary factor $c$ with the length-scale $\ell$ of the approximated function. In this section we discuss about the existing theoretical evidence that the minimum number of basis functions $m$ needed for an accurate approximation is linearly proportional to the the boundary factor $c$ and inversely proportional to the length-scale $l$.
% In the next Section \ref{subsec_empiric_relations} we conduct an empirical analysis through simulations to prove this expected linear proportionality between $m$, $c$ and $l$ and compute the specific coefficients of proportionality for every kernel.

% {\color{blue}Aki: We can also look at the mean zero-level upcrossings, which for QE is
% proportional to 1/l which gives as that part, and the effect of c should also
% be quite linear when c is not very close to 1 and m is not very small as it
% linearly makes the interval shorter so that the there are less zero-level
% upcrossings available in the basis functions. You can visualize this using
% {\small \url{https://avehtari.github.io/casestudies/Motorcycle/motorcycle_gpcourse.html}} and
% look how the plot of 6 basis functions change when you set c=1, c=1.5, c=2.25
% (ie $1.5^2$). Instead of looking at mean zero-level crossings, we can look at the
% spectral densities and when they go to zero. This is illustrated in the same
% notebook after that figure. E.g. with length-scale l=1,l=1/2, l=1/4, l=1/8 to
% explain 99\% of the prior variance we need 4, 8, 16, and 32 basis functions
% (linear in l) and to explain 80\% of the prior variance we need 2, 4, 8, 16
% basis functions. These spectral densities don't depend on c. When c increaes
% the basis functions are truncated and for example going from c=1.5 to c=2.25,
% basis function 9 has the flexibility basis function 6 had, so 1.5 more basis
% functions are needed. So we have proportionality c/l. }

@@

\subsection{Empirical discovering of the functional form of the relationships between $m$, $c$ and $\ell$} \label{subsec_empiric_relations}

\begin{figure*}
\centering
\subfigure{\includegraphics[width=\textwidth, trim = 0mm 0mm 0mm 0mm, clip]{fig6_relationships.png}}
\caption{Relation among the minimum number of basis functions $m$, the boundary factor $c$ \hspace{0.1mm} ($c = \frac{L}{S}$) \hspace{0.1mm} and the length-scale normalized by the half-range of the data ($\frac{\ell}{S}$), for the squared exponential, Mat\'ern ($\nu$=3/2) and Mat\'ern ($\nu$=5/2) covariance functions.}
  \label{fig6_relationships}
\end{figure*}

Empirical simulations are carried out to analyze the relationships between $m$, $c$ and $l$. Figure~\ref{fig6_relationships} depicts how these three factors interact and affect the accuracy of the HSGP approximation for a GP with squared exponential, Mat\'ern ($\nu$=3/2) and Mat\'ern ($\nu$=5/2) covariance functions and a single input dimension. More precisely, for a given GP model (with a particular covariance function) with length-scale $\ell$ and given a boundary factor $c$, Figure~\ref{fig6_relationships} shows the minimum number of basis functions $m$ required to obtain an accurate approximation in the sense of satisfying eq.~\eqref{eq_diff_covs}. We considered an approximation to be a sufficiently accurate when the total variation difference between the approximate and exact covariance functions, $\varepsilon$ in eq.~\eqref{eq_diff_covs}, is below 1$\%$ of the total area under the curve of the exact covariance function $k$ such that
%
\begin{equation} \label{eq:rel_total_distance}
 \int | k(\tau) - \tilde{k}_m(\tau)|  \,\mathrm{d}\tau  < 0.01 \int k(\tau) \,\mathrm{d}\tau,
\end{equation}
%
where $\tilde{k}_m$ is the approximate covariance function with $m$ basis functions. Alternatively, these figures can be understood as providing the minimum $c$ that we should use for given $\ell$ and $m$. Of course, we may also read it as providing the minimum $\ell$ that can be approximated with high accuracy given $m$ and $c$. We obtain the following main conclusions:
%
\begin{itemize}
\item As $\ell$ increases, $m$ required for an accurate approximation decrease.
\item The lower $c$, the smaller $m$ can and $\ell$ must be to achieve an accurate approximation.
\item For a given $\ell$ there exist a minimum $c$ under which a accurate approximation is never going to be achieved regardless of $m$. This fact can be seen in Figure~\ref{fig6_relationships} as the contour lines which represent $c$ have an end in function of $\ell$ (Valid $c$ are restricted in function of $\ell$). As $\ell$ increases, the minimum valid $c$ also increases.
\end{itemize}

@@

As explained in Section~\ref{subsec_theoretic_linearity}, when $c$ is large enough, there is a near linear proportionality between $m$, $\frac{l}{S}$ (length-scale $l$ relative to the half range of the input domain $S$) and $c$. To obtain practical numerical functions that can be used to guide the selection of these parameters, we have empirically derived the practically useful constant terms. We require a lower bound for $c$ of $c \geq 1.2$ such that the equations below are precise enough for practical application.

@@

\paragraph*{Squared exponential}:
%
\begin{align}\label{eq_m_l_QE}
&m= 1.75 \, \frac{c}{\nicefrac{l}{S}} \;\; \Leftrightarrow \;\; \nicefrac{l}{S}= 1.75 \, \frac{c}{m},
\end{align}
%
with
%
\begin{align}\label{eq_c_vs_l_QE}
&c \geq \, 3.2 \, \nicefrac{l}{S} \;\; \& \;\, c \geq \, 1.2 .
\end{align}

\paragraph*{Mat\'ern ($\nu$=5/2)}:
%
\begin{align}\label{eq_m_l_mat52}
&m= 2.65 \, \frac{c}{\nicefrac{l}{S}} \;\; \Leftrightarrow \;\; \nicefrac{l}{S}= 2.65 \, \frac{c}{m},
\end{align}
%
with
%
\begin{align}\label{eq_c_vs_l_mat52}
&c \, \geq \, 4.1 \, \nicefrac{l}{S} \;\; \& \;\, c \geq \, 1.2 .
\end{align}

\paragraph*{Mat\'ern ($\nu$=3/2)}:
%
\begin{align}\label{eq_m_l_mat32}
&m= 3.42 \, \frac{c}{\nicefrac{l}{S}} \;\; \Leftrightarrow \;\; \nicefrac{l}{S}= 3.42 \, \frac{c}{m},
\end{align}
%
with
%
\begin{align}\label{eq_c_vs_l_mat32}
&c \, \geq \, 4.5 \, \nicefrac{l}{S} \;\; \& \;\, c \geq \, 1.2 .       
\end{align}

These constants vary monotonically with respect to $\nu$ (squared exponential corresponding to Mat\'ern with $\nu \to \infty$). Using the formula for Mat\'ern ($\nu$=3/2) provides the largest $m$ and $c$, and thus this formula alone could be used as a conservative choice for all Mat\'ern covariance functions with $\nu \geq 3/2$ and likely as a good initial guess for many other covariance functions. If the aim is to find minimal $m$ to speedup the computation, a further refined formula can be obtained for new covariance functions.

Figure~\ref{fig6_relationships} and previous equations~(\ref{eq_m_l_QE})-(\ref{eq_c_vs_l_mat52}) were obtained for a GP with a unidimensional covariance function, which result in a surfaces depending on three variables, $m$, $c$ and $\ell$. Equivalent results for a GP model with a two-dimensional covariance function would result in a surface depending on four variables, $m$, $c$, $\ell_1$ and $\ell_2$. More precisely, in the multi-dimensional case, whether the approximation is close enough might depend only on the ratio between wigglyness in every dimensions. For instance, in the two-dimensional case, it would depend on the ratio between $\ell_1$ and $\ell_2$. Future research will focus on building useful graphs or analytical models that provide these relations in multi-dimensional cases. However, as an approximation, we can use the unidimensional GP conclusions in Figure~\ref{fig6_relationships} or equations~(\ref{eq_m_l_QE})-(\ref{eq_c_vs_l_mat52}) to check the accuracy by analyzing individually the different dimensions of a multidimensional GP model.

@@

The numerical equation governing the relationships between $m$ and $\ell$ for a periodic squared exponential covariance function was estimated in eq.~\eqref{eq_J_l_periodic} which is depicted next:
%
\begin{align*}
&m \geq \frac{3.72}{\ell} \;\; \Leftrightarrow \;\; \ell \geq \frac{3.72}{m}
\end{align*}

@@

Guidelines for updating $m$ and/or $c$ are:
%
\begin{itemize}
	\item $c$ must be big enough for the given length-scale $l$ of the function to be learned and,
	\item $m$ must be big enough for the given $\ell$ and $c$. 
\end{itemize}

@@

If the initial length-scale is not far from the optimal, and larger than minimal $c$ and $m$ are used in the initial HSGP model, it is likely that the initial results are already sufficiently accurate. The following iterative procedure works as safety approach and also allows finding a smaller sufficient $m$ which can be useful if the user wants to optimize the computational speed for repeated fitting (e.g., in cross-validation or simulation based calibration).

@@

Based on the above proposed diagnostics, we obtain a simple, iterative step-by-step procedure that users can follow in order to obtain an accurate HSGP approximation. The procedure is split into two phases, Phase~A and B, which have to be completed consecutively.
%In the following, we will refer to the Equations from Section XXX for the double-exponential kernel, but the same procedure works for other kernels by using the corresponding Equations provided in the same Section.

@@

\item[A1.] Make an initial guess on the length-scale $\ell_1$. If there is no useful information available, we recommend to start with a large length-scale, about $\ell_1 \in [0.5, 1]$. Set the iteration index to $i=1$.

\item[A2.] Obtain the minimum valid boundary factor $c_i$ determined by $\ell = \ell_i$ for the given kernel as per Section~\ref{sec_num_equations}.

\item[A3.] Obtain the mimimum valid number of basis functions $m_i$ determined by $\ell = \ell_i$ and $c = c_i$ for the given kernel as per Section~\ref{sec_num_equations}.

@@

\item[A5.] Perform the length-scale diagnostic by checking if $\hat{\ell}_i - 0.01 \geq \ell_i$.

@@

If the diagnostic is TRUE, the HSGP approximation is close to be sufficiently accurate. Continue with Phase~B.

@@

\item[B1.] For the current HSGP model, compute measures of predictive accuracy, for example, root mean square error (RMSE), coefficient of determination ($R^2$), or expected log predictive density (ELPD).

@@

\item[B4.] Obtain the minimum valid length-scale $\ell_i$ that can be accurately estimated by $m = m_i$ and $c = c_i$ for the given kernel as per Section~\ref{sec_num_equations}.

@@

\item[B6.] Perform the length-scale diagnostic by checking if $\hat{\ell}_i - 0.01 \geq \ell_i$.

@@

If all the stability checks succeed, the HSGP approximation of the latest model should be sufficiently accurate and the procedure ends here.

@@

\begin{figure*}
\centering
\subfigure{\includegraphics[width=\textwidth, trim = 0mm 0mm 0mm 0mm, clip]{fig7_lscale_comparison.png}}
\caption{Exact GP length-scale ($\ell_{GP}$) and approximate HSGP length-scale ($\hat{\ell}$) for various datasets with varying smoothness (left) and their corresponding root mean squared errors (RMSE) computed with the exact GP model as the reference.}
  \label{fig7_lscale_comparison}
\end{figure*}

@@

In Section \ref{sec_accuracy}, we thoroughly studied the relationship between the number of basis functions in the approximation and the approximation accuracy across different configurations. We specifically focused on the Matérn and squared exponential families of covariance functions, yet there exists other families of stationary covariance functions. The basis function approximation can easily be implemented for any stationary covariance function, where the spectral density function is available. The assess the accuracy of a basis function approximation for a kernel, where the diagnostic plots like Figure \ref{fig6_relationships} or equations (\ref{eq_m_l_QE})-(\ref{eq_c_vs_l_mat52}) are not available, we suggest to use the relative total variational distance between the true covariance function and the approximate covariance function as given in eq. \eqref{eq:rel_total_distance}. Ensuring that the relative distance is bounded by a small constant for the relevant length-scale implies a high quality approximation.

@@

A GP model with a periodic covariance function does no fit in the framework of the HSGP approximation covered in this study, but it has also a low-rank representation. In this section, we first give a brief presentation of the results by \citet{solin2014explicit}, who obtain an approximate linear representation of a periodic squared exponential covariance function based on expanding the periodic covariance function into a series of stochastic resonators. Secondly, we analyze the accuracy of this approximation and, finally, we derive the GP model with this approximate periodic squared exponential covariance function.
